The abundance of Bullet-Groups in $\Lambda$CDM

### 2014-03-11 paper/ms.tex
figure
new intro with references

\newcommand{\avg}[1]{\langle{#1}\rangle}  
\newcommand{\hMpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hGpc}{{\ifmmode{h^{-1}{\rm Gpc}}\else{$h^{-1}$Gpc }\fi}}  
\newcommand{\hmpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hkpc}{{\ifmmode{h^{-1}{\rm kpc}}\else{$h^{-1}$kpc }\fi}}  
\newcommand{\hMsun}{{\ifmmode{h^{-1}{\rm {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}  
\newcommand{\hmsun}{{\ifmmode{h^{-1}{\rm {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}  
\newcommand{\Msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\kms}{{\ifmmode{{\mathrm{\,km\ s}^{-1}}}\else{\,km~s$^{-1}$}\fi}}
\newcommand{\bullb}{MACS J0025.4-1222}
\newcommand{\bulla}{1E0657---56} 
\newcommand{\bullg}{SL2S J08544-0121}
\shorttitle{Bullet Groups}
\shortauthors{Fern\'andez-Trincado et al.}

@@

\title{The abundance of Bullet Groups in $\Lambda$CDM}
\author{J. G. Fern\'andez-Trincado$^{1,2,3}$, J. E. Forero-Romero$^1$,
  T. Verdugo$^3$, G. Foex$^4$ and V. Motta$^4$} 
\affil{$^1$ Departamento de F\'{i}sica, Universidad de los Andes,
  Cra. 1 No. 18A-10, Edificio Ip, Bogot\'a, Colombia\\ 
       $^2$ Institute Utinam, CNRS UMR6213, Universit\'e de
  Franche-Comt\'e, OSU THETA de Franche-Comt\'e-Bourgogne,
  Besan\c{c}on, France\\ 
       $^3$ Centro de Investigaciones de Astronom\'ia, AP 264,
  M\'erida 5101-A, Venezuela\\
  $^4$ Departamento de F\'isica y Astronom\'ia, Universidad de
  Valpara\'iso, Avda. Gran Breta\~na 1111, Playa Ancha, Valpara\'iso
  2360102, Chile  
}

@@

We estimate the expected distribution of displacements between the two
dominant dark matter peaks in two groups of halos corresponding to
galaxy groups and clusters. We find that 50\% to 60\% of the dark
matter  with circular velocities in the range $300\kms$ to $700\kms$
(groups) show multimodal morphologies with displacements between the
dark matter clumps equal or larger than $133\pm21$\hkpc, which
corresponds to the observational constraint the object SL2S
J08544-0121 located at $z=0.35$ which has a velocity dispersion of
$644^{+69}_{-102}\kms$. For dark matter halos with circular velocities
larger than $700\kms$ (clusters) this fraction rises to $80\%$ to
$85\%$. Using the same simulation we estimate the dark
matter-to-baryon spatial separation and find that that $0.1\%$ to
$1.0\%$ of the low mass systems should present separations equal or
larger than $87\pm 14$\hkpc corresponding to the same observational
benchmark; for halos with circular velocities larger than $700\kms$
this fraction is in the range of $4\%$ to $10\%$, consistent with
previous studies of dark matter to baryon separations in
clusters. These results present a weak dependence with redshift. 
\end{abstract}

@@

The Bullet Cluster provided a new kind of observational evidence of
the existence of dark matter. Quantifying the displacement between dark
matter and the dominant baryonic component (hot X-ray emitting gas)
\citep{ForeroRomero2010} has been used to test the CDM paradigm itself
by quantifying the substructure velocity required to produce such
displacement \citep{Hayashi2006} to finally estimate the expected
abundance of such events in the Universe.

Since that time there are other observations of other Bullet-like
systems [...].

Recently \citep{Gastaldello} observed baryonic-DM displacement if
$124\pm 20$ kpc in a group-like system with a total mass $2.4\pm 0.6
\times 10^{14}$\Msun. Systems of this mass are $\sim 10$ times more massive
than cluster systems in the mass range $>10^{15}$\hMsun, this opens
up the possibility of observationally finding bullet groups in a fair
amount to impose constraints on $\Lambda$CDM. This greater abundance
has to be weighted by the fraction of systems that present large
displacements. Such study has been performed for clusters but not for
lower mass systems.

In this Letter we present prediction for the abundance of group-like
systems that might show a DM-baryon displacement. To this end we use a
N-body cosmological simulation with such a resolution that allows us
to identify multi-modal dark matter clumps in the circular velocity
range $300-700$\kms.

This paper is organized as the follows. In Section
\ref{sec:simulation} we present the simulation and the halo catalogs
used in this work. We continue in Section \ref{sec:setup} with the
geometry of the problem at hand and the measurements setup. Next in
Section \ref{sec:results} we present our results to finish with a
discussion and conclusion in Sections \ref{sec:discussion} and
\ref{sec:conclussions}.

@@

We use the Bolshoi Run, a cosmological DM only simulation over a cubic
volume of 250 comoving \hMpc on a side. The simulation uses the ART code to
follow the evolution of  a dark matter density field sampled with
$2024^3$ from $z=80$ to $z=0$ [...]. The cosmology used  corresponds
to  the spatially flat concordance model with the following
parameters:  the density parameter for matter (dark matter$+$baryons)
$\Omega_m=0.27$, the density parameter for baryonic matter
$\Omega_b=0.0469$, the density parameter for dark energy
$\Omega_{\Lambda}=0.73$, the Hubble parameter $h=0.7$, the
normalization of the Power spectrum $n=0.95$ and the amplitude of mass
density fluctuation (at redshift z$=$0) $\sigma_8=0.82$.  The number
of particles used for each of the DM component was $2048^3$, resulting
in a mass resolution of $1.35 \times 10^8$
M$_{\odot}$h$^{-1}$. \citet{2011ApJ...740..102K}.

@@

Figure \ref{fig:} presents the geometry of the bullet groups using the 
variables $\mu$ and $D_{\rm off}$. The first evident feature is that
most of the configurations have $|mu|>0.9$ ($\theta\leq 30^{\circ}$),
meaning that most of the collisions can be described as a head-on while
only a minority with $|mu|<0.9$ have more tangential trajectories. For
the pairs on radial trajectories there are three regions of interest
in this plane that describe different stages in the collision,
assuming that the sub-halo merges or falls below the detection
threshold after the first pass through the center of the host halo.

>>>>

\newcommand{\avg}[1]{\langle{#1}\rangle}  
\newcommand{\hMpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hGpc}{{\ifmmode{h^{-1}{\rm Gpc}}\else{$h^{-1}$Gpc }\fi}}  
\newcommand{\hmpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hkpc}{{\ifmmode{h^{-1}{\rm kpc}}\else{$h^{-1}$kpc }\fi}}  
\newcommand{\hMsun}{{\ifmmode{h^{-1}{\rm {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}  
\newcommand{\hmsun}{{\ifmmode{h^{-1}{\rm {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}  
\newcommand{\Msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\kms}{{\ifmmode{{\mathrm{\,km\ s}^{-1}}}\else{\,km~s$^{-1}$}\fi}}
\newcommand{\bullb}{MACS J0025.4-1222}
\newcommand{\bulla}{1E0657---56} 
\newcommand{\bullg}{SL2S J08544-0121~}
\shorttitle{Bullet Groups}
\shortauthors{Fern\'andez-Trincado et al.}

@@

\title{The abundance of Bullet-Groups in $\Lambda$CDM}
\author{J. G. Fern\'andez-Trincado$^{1,2,3}$, J. E. Forero-Romero$^1$,
  T. Verdugo$^3$, G. Foex$^4$ and V. Motta$^4$} 
\affil{$^1$ Departamento de F\'{i}sica, Universidad de los Andes,
  Cra. 1 No. 18A-10, Edificio Ip, Bogot\'a, Colombia\\ 
       $^2$ Institute Utinam, CNRS UMR6213, Universit\'e de
  Franche-Comt\'e, OSU THETA de Franche-Comt\'e-Bourgogne,
  Besan\c{c}on, France\\ 
       $^3$ Centro de Investigaciones de Astronom\'ia, AP 264,
  M\'erida 5101-A, Venezuela\\
  $^4$ Departamento de F\'isica y Astronom\'ia, Universidad de
  Valpara\'iso, Avda. Gran Breta\~na 1111, Playa Ancha, Valpara\'iso
  2360102, Chile  
}

@@

We estimate the expected distribution of displacements between the two
dominant density peaks in dark matter halos simulated in a full
cosmological context. We use as a benchmark the
observation of \bullg, which is the lowest mass system observed so far
featuring a bi-modal dark matter distribution with a dislocated 
baryonic component. The study extends previous theoretical studies by
simulating two samples: groups and clusters. We find that 50\% to 60\%
of the dark matter halos with circular velocities in the range
$300\kms$ to $700\kms$ (groups) show multimodal morphologies with
displacements between the dark matter clumps equal or larger than
$133\pm21$\hkpc observed in \bullg. For dark matter halos with
circular velocities larger than $700\kms$ (clusters) this fraction
rises to $80\%$ to $85\%$. Using the same simulation we estimate the
dark matter-to-baryon spatial separation and find that $0.1\%$ to
$1.0\%$ of the low mass systems should present separations equal or
larger than $87\pm 14$\hkpc corresponding to our observational
benchmark; for halos with circular velocities larger than $700\kms$
this fraction is in the range of $4\%$ to $10\%$, consistent with
previous studies of dark matter to baryon separations in clusters. The
extension of the theoretical predictions and observational results
towards low mass Bullet-like configurations, i.e. larger abundance of
systems,  opens up the possibility for a new statistical test of
$\Lambda$CDM.

\end{abstract}

@@

The Bullet Cluster (massive cluster of galaxies \bulla) provided a new
kind of observational evidence for the existence of dark matter
\citep{Markevitch2004,Clowe2006}. Quantifying the displacement between
dark matter and the dominant baryonic component (hot X-ray emitting gas)
has been used to test the Cold Dark Matter (CDM)
paradigm itself by quantifying the substructure velocity required to
produce such displacement \citep{Hayashi2006, Springel2007,
  Thompson2012}, the displacemeent between the dominant dark matter
and baryonic component \citep{ForeroRomero2010} to estimate the
expected abundance of such events in a $\Lambda$CDM Universe and even
explore possible extensions to the concordance cosmological model
\citep{Farrar2007,Lee2010,Lee2012}.

Since then, other examples of Bullet-like systems have been found; 
MACS J0025.4-1222 \citep{Bradac2008}, Abell 2744 \citep{Merten2011},
DLSCL J0916.2+2951 \citep{Dawson2012}, ZwCl 1234.0+02916
\citep{Dahle2013}. Recently \citep{Gastaldello} observed baryonic-DM
displacement if $124\pm 20$ kpc in a group-like system with a total
mass $2.4\pm 0.6 \times 10^{14}$\Msun. Systems of this mass are
$\sim10$ times more abundant than cluster systems in the mass range of
the Bullet Cluster $>10^{15}$\hMsun, this opens up the possibility of
observationally finding bullet groups in a fair amount to impose
constraints on $\Lambda$CDM.

However, a greater abundance of small mass systems has to be weighted
by the probability of having a merger and 
presenting a large displacement between the DM and baryonic
components. These two conditions (merger rates, maximum possible
displacement) are a funcion of halo mass in $\Lambda$CDM
cosmologies.  Such study has been performed for clusters but not for
lower mass systems \citep{ForeroRomero2010}. The existence of objects
like \bullg and observational campaigns like the Strong Lensing Legacy
Survey (SL2S) open up the possibility of finding multi-modal dark
matter distributions in the group mass range.

In this Letter we compute the abundance of group-like dark
matter halos with a multi-modal morphology that also might present a
DM-baryon displacement. To this end we use a N-body cosmological
simulation with such a resolution that allows us to identify
multi-modal dark matter distributions in hosts with circular velocities
larger than $300$\kms.

This paper is organized as the follows. In Section
\ref{sec:simulation} we present the simulation and the halo catalogs
used in this work. We continue in Section \ref{sec:setup} with the
geometry of the problem at hand and the measurements setup. Next in
Section \ref{sec:results} we present our results and observational
perspectives to finally conclude in Sections \ref{sec:conclusions}.

@@

We use the Bolshoi Run, a cosmological DM only simulation over a cubic
volume of 250 comoving \hMpc on a side. The simulation uses the ART code to
follow the evolution of  a dark matter density field sampled with
$2048^3$ from $z=80$ to $z=0$ [...]. The cosmology used  corresponds
to  the spatially flat concordance model with the following
parameters:  the density parameter for matter (dark matter$+$baryons)
$\Omega_m=0.27$, the density parameter for baryonic matter
$\Omega_b=0.0469$, the density parameter for dark energy
$\Omega_{\Lambda}=0.73$, the Hubble parameter $h=0.7$, the
normalization of the Power spectrum $n=0.95$ and the amplitude of mass
density fluctuation (at redshift z$=$0) $\sigma_8=0.82$.  The number
of particles used for each of the DM component was $2048^3$, resulting
in a mass resolution of $1.35 \times 10^8$
M$_{\odot}$h$^{-1}$. \citet{2011ApJ...740..102K}.

@@

Figure \ref{fig:geometry} presents the geometry of the bullet groups using the 
variables $\mu$ and $D_{\rm off}$. The first evident feature is that
most of the configurations have $|mu|>0.9$ ($\theta\leq 30^{\circ}$),
meaning that most of the collisions can be described as a head-on while
only a minority with $|mu|<0.9$ have more tangential trajectories. For
the pairs on radial trajectories there are three regions of interest
in this plane that describe different stages in the collision,
assuming that the sub-halo merges or falls below the detection
threshold after the first pass through the center of the host halo.
